Stage-Splitting a Modal Language

### 2014-06-15 paper/semantics.tex
splitting up figures into separate files
Clarifying the semantics.  Still more needs to be done.

In this section, we discuss the dynamic semantics of our language.  Before we can get into the details of the semantics, we have to first make some decisions about the behavior we want it to have.  Similar to \cite{davies96}, we want our semantics to have the form of a judgement that partially evaluates an expression to create a residual that can be evaluated at the second stage in the standard way.  Again like Davies, we choose a call-by-value paradigm as our baseline at both stages.  The only significant freedom that this leaves is how to handle $\next$, $\prev$, and $\pause$ constructs.

There are two feasible ways to partially evaluate the following stage-\bbtwo~expression:
\begin{lstlisting} 
let x = (next {4+5}, 7+8) in
next{
	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
}
\end{lstlisting}
We can either duplicate the contents of the first $\next$ expression to produce,
\begin{lstlisting}
(4+5) * (4+5) * 15
\end{lstlisting}
or we can bind the contents of the $\next$ expression, and simply duplicate the variable,
\begin{lstlisting} 
let x1 = 4+5 in
x1 * x1 * 15
\end{lstlisting}

Davies chose the first, whereas we choose the second technique. We think it makes more sense from the point of view of cost.  That said, it's not immediately clear how to get this behavior.  We first present an informal example to give an intuition for what our semantics does.  We start with the same example as before, but wrapped with a $\prev$:
\begin{lstlisting}
prev {
let x = (next {4+5}, 7+8) in
next{
	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
}
}
\end{lstlisting}
The first thing we need to do is evaluate the binding for $x$.  Of course, we can't evaluate inside the $\next$ statement, but we still need to produce a value that will get substituted.  What we do instead is replace the body of the $\next$ by a variable and make an explicit substitution, which then floats up to the top of the containing $\prev$:
\begin{lstlisting} 
prev{
[4+5/x1]
let x = (next{x1}, 15) in
next{
	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
}
}
\end{lstlisting}
We're now free to perform the stage-\bbone~substitution for $x$.  
\begin{lstlisting} 
prev{
[4+5/x1]
next{
	prev{#1 (next{x1}, 15)} 
	* prev{#1 (next{x1}, 15)} 
	* prev{hold (#2 (next{x1}, 15))}
}
}
\end{lstlisting}
To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
\begin{lstlisting} 
prev{
[4+5/x1]
	next{ x1 * x1 * 15 }
}
\end{lstlisting}
The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
\begin{lstlisting} 
let x1 = 4+5 in
 x1 * x1 * 15
\end{lstlisting}

>>>>

In this section, we discuss the dynamic semantics of our language.  
Before we can get into the details of the semantics, we have to first make some decisions about the behavior we want it to have.  
Similar to \cite{davies96}, we want our semantics to have the form of a judgement that partially evaluates an expression to create a residual that can be evaluated at the second stage in the standard way.  Again like Davies, we choose a call-by-value paradigm as our baseline at both stages.  The only significant freedom that this leaves is how to handle the staging constructs ($\next$, $\prev$, and $\pause$).

We present here two reasonable interpretations of these staging constructs, 
in the context of the following stage-\bbtwo~example:
\begin{lstlisting} 
prev{
let x = (next {4+5}, 7+8) in
next{
	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
}
}
\end{lstlisting}
The first option is to duplicate the contents of the first $\next$ expression to produce,
\begin{lstlisting}
(4+5) * (4+5) * 15
\end{lstlisting}
Essentially, we're treating $\next$ as a value a stage-\bbone, so long as all of its contained $\prev$s are reduced.  
This is the technique of \cite{davies96}.
Instead of this, we take a second, nearly equivalent approach.  
That is to bind the contents of the first $\next$ expression to some variable (here, $x1$) and then duplicate that variable,
\begin{lstlisting} 
let x1 = 4+5 in
x1 * x1 * 15
\end{lstlisting}

We think that this interpretation makes more sense from the point of view of cost, 
since the contents of a $\next$ always appear once in the residual, regardless of how many times the result is used.  
This is {\em nearly} equivalent to the first expression, because they have different termination behavior 
when a non-terminating stage-\bbtwo~expression is declared but never used.

Now that we know {\em what} the behavior is that we want, we can explain {\em how} our semantics achieves that behavior.  We start with the same example:
\begin{lstlisting}
prev {
let x = (next {4+5}, 7+8) in
next{
	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
}
}
\end{lstlisting}
The first thing we need to do is evaluate the binding for $x$.  Of course, we can't evaluate inside the $\next$ statement, but we still need to produce a value that will get substituted.  What we do instead is replace the body of the $\next$ by a variable and make an explicit substitution, which then floats up to the top of the containing $\prev$:
\begin{lstlisting} 
prev{
[4+5/x1]
let x = (next{x1}, 15) in
next{
	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
}
}
\end{lstlisting}
We're now free to perform the stage-\bbone~substitution for $x$.  
\begin{lstlisting} 
prev{
[4+5/x1]
next{
	prev{#1 (next{x1}, 15)} 
	* prev{#1 (next{x1}, 15)} 
	* prev{hold (#2 (next{x1}, 15))}
}
}
\end{lstlisting}
To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
\begin{lstlisting} 
prev{
[4+5/x1]
	next{ x1 * x1 * 15 }
}
\end{lstlisting}
The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
\begin{lstlisting} 
let x1 = 4+5 in
 x1 * x1 * 15
\end{lstlisting}

Once again, the point here is that we did not duplicate the $4+5$ in our residual.  
Instead we lifted it out into a substitution, leaving a hatted variable [note: add hats above] in its place,
and we reified the substitution into a let-binding at the enclosing $\prev$.

\subsection{more formally...}
